---
title: "BIO720_2021_FunctionsAssignment_Answers"
author: "ID"
date: "16/11/2021"
output:
  html_document:
    toc: yes
    number_sections: yes
    keep_md: yes
  pdf_document:
    toc: yes
editor_options:
  chunk_output_type: console
---




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 3)
```


## When marking the questions


*Please note that many of the questions were purposefully open ended in how they could be answered. i.e different people might write the functions slightly differently, or set warnings differently. So it may be useful to look at how they answered the question, and try different kinds of inputs to see how it handles potential problems.*

*If you are not sure how to grade a particular question please contact me.*

1. DataCamp course "Intermediate R" Chapters 1- 4. Send me a picture showing you have completed the appropriate chapters, but I should be able to see the completion status on DataCamp as well.

*2 marks*

## R functions to write


2. Write a named function that can take a single numeric vector as an input (argument), and outputs (as a *numeric vector*) the sample size, arithmetic mean, standard deviation, standard error of the mean and the coefficient of variation. Make sure to include in the function ways of dealing with unexpected input values like **a)** missing data in the input vector, **b)** an empty input vector, **c)** non-numeric values in the input vector and **d)** provides a warning if the sample size is too small in the input vector (a sample size of 5 or less). Make sure it computes the *correct sign (+)* for the coefficient of variation. Finally make sure the output of this function is clear (i.e. we know which elements of the vector correspond to the mean, standard deviation, etc).

*Please note that my function is just one way of handling it.

```{r}
function2 <- function(x) {
  
  if (mode(x) != "numeric") stop("x is not numeric") # function fails if values are not numeric
  if (anyNA(x)) warning("missing data present, will be omitted")
  
  x <- na.omit(x)
  
  if (length(x) == 0) stop("x is an empty vector")
  else if (length(x) < 6) warning("sample size is very small, estimates will be unreliable")

  
  ss_x <- length(x) # sample size
  mean_x <- mean(x) # mean
  sd_x <- sd(x)     # standard deviation
  se_x <- sd(x)/sqrt(ss_x)  # standard error of the mean
  cv_x <- sd_x/abs(mean_x)  # note abs to keep CV +ve
  
    # returned numeric vector. List would be fine too.
  return(c(sample_size = ss_x,
           mean = mean_x,
           stdDev = sd_x,
           stdErr = se_x,
           CV = cv_x))
}
```


3. Write a named function that can take two named input numeric vectors as arguments. In the body of the function compute the element by element ratios from the two vectors, and then computes the arithmetic mean of those ratios. This function also should compute the ratio of the means of the two vectors. Then, as a list output an object with the following elements: **a** the mean of the element-by-element ratios, **b** the ratio of the means and **c** the difference between these two values. Given the goals of this function (computing these three elements), use the ideas we discussed in class and what you used in the previous function to build in appropriate "safety checks" and warnings against unexpected inputs, or inputs that would result in erroneous output.


```{r}
ratio_function <- function(x, y) {
  
    # If there are a lot of conditions, this can be useful (but a bit more cryptic for error messages)
    stopifnot(exprs = {
      length(x) == length(y)  # make sure vectors are the same length
      mode(x) == "numeric"    # make sure x is numeric
      mode(y) == "numeric"})
  
    if (anyNA(x) || anyNA(y)) warning("missing data present, will be omitted element-wise")
  
    ratio_values <- na.omit(x/y) # in case there are any element specific missing data
    mean_of_ratios <- mean(ratio_values)
    ratio_of_means <- mean(x, na.rm = T)/mean(y, na.rm = T)
    
    ratio_difference <- mean_of_ratios - ratio_of_means
    
    return(list(mean_of_ratios = mean_of_ratios, 
                   ratio_of_means = ratio_of_means,
                   difference = ratio_difference))
    }

```


4. Now make a revised version of the function you wrote for question 3, where the user of the function can choose to use either the mean or the median (but the functions default behaviour is the mean like in the function you wrote for question 3).

*In this case I just used an if else to separate the expressions using either median or mean*

```{r}
ratio_function_v2 <- function(x, y, med = F) {
  
    # If there are a lot of conditions, stopifnot can be useful (but a bit more cryptic error messages)
    stopifnot(exprs = {
      length(x) == length(y)  # make sure vectors are the same length
      mode(x) == "numeric"    # make sure x is numeric
      mode(y) == "numeric"})
  
    if (anyNA(x) || anyNA(y)) warning("missing data present, will be omitted element-wise")
  
    ratio_values <- na.omit(x/y) # in case there are any element specific missing data
    
    if (med == T) {
    
      med_of_ratios <- median(ratio_values)
      ratio_of_medians <- median(x, na.rm = T)/median(y, na.rm = T)
      ratio_difference <- med_of_ratios  - ratio_of_medians
    
      return(list(median_of_ratios = mean_of_ratios, 
                   ratio_of_medians = ratio_of_means,
                   difference = ratio_difference))}
    
    else {
      mean_of_ratios <- mean(ratio_values)
      ratio_of_means <- mean(x, na.rm = T)/mean(y, na.rm = T)
      ratio_difference <- mean_of_ratios - ratio_of_means
    
      return(list(mean_of_ratios = mean_of_ratios, 
                   ratio_of_means = ratio_of_means,
                   difference = ratio_difference))}
    }

```


5.  Read in the sequence read below (from an Illumina HiSeq run). Write a function that can take a sequence read like this as input and outputs a table of the frequencies of A, C, G & T (two digits for each) along with the number of base pairs in the read. Keep in mind a few things. You may wish to split the read from a single string into a vector where each element is a single base pair (or into a list if you prefer). There are numerous options but one approach would be to use the `strsplit` function. This will output a list (not a character vector). If you want to use the length function note that it won't have the expected behaviour on a list (but if you look at the help for length you will see an option for this). Some other functions that *may* help you on your way (use R help to understand how they can be used) include: `table`, `nchar`, `length`, `lengths`, `print`, `unlist`, `mode`.



```{r}
read_1 <- "CGCGCAGTAGGGCACATGCCAGGTGTCCGCCACTTGGTGGGCACACAGCCGATGACGAACGGGCTCCTTGACTATAATCTGACCCGTTTGCGTTTGGGTGACCAGGGAGAACTGGTGCTCCTGC"
```

*This serves as an answer for both 5 and 6.*

```{r}

BaseFrequencies <- function(x) {
    
    # if it is a single string still
    if (length(x) == 1 & mode(x) == "character") {
    	x <- strsplit(x, split = "", fixed = T) 
        x <- as.character(unlist(x))
    }     
    
    if (mode(x) == "list") {
    	tab <- table(x)/lengths(x)
      seq_len <- lengths(x)}
  
    else {
    	tab <- table(x)/length(x)
    	seq_len <- length(x)
    }
   #return(c(tab, sequence_length = seq_len))  # as a vector
   return(list(seq_frequencies = tab, sequence_length = seq_len)) # as a list
}

```


6. Can you write a revised version of your function from question 5  so that it can handle the input of either a string as a single multi-character vector (like `read_1`), **or** a character vector of individual nucelotide bases **or** a list with each element being a character for an individual base (like you would get from using `strsplit()` on `read_1`)? The `mode` function may be useful for this question (depending on how you choose to code it).

7. While you don't need to write the actual changes to the function from question 6 (but feel free to do so), provide at least three things you might build checks or warnings for that given the function, and possibly "incorrect" inputs. i.e some things a user of the function may try to use as input that would result in unexpected or incorrect outputs.

*- handling lower and upper case nucleotides (lower case usually signifies a base of low quality*
*- handling other IUPAC nucleotide characters. For instance, what do you want to do with N's in the sequence? Count or remove*
*- error messages if nonsensical characters (like numbers) are included*

