---
title: "Bio720_MuckingWithDrosophilaPromoters"
author: "Ian Dworkin"
date: "12/3/2018"
output: html_document
---

# A little introduction to Bioconductor using Drosophila


## A visit to Bioconductor
You have to give the developers of Bioconductor a little love. They have really gone out of their way to provide very powerful *R* libraries, with lots of tutorials. Let's take a look shall we. The bioconductor home page is [here](http://bioconductor.org/).

## A little introduction with Drosophila

With 1649 packages it is not like I can show you everything. However your book, and the datacamp tutorial has given you a tiny bit of an intro into things.

We will just use a few packages today

```{r}
library(Biostrings)
library(BiocInstaller) # old deprecated
library(BiocManager) # new way
library(GenomicRanges)
library(GenomicFeatures)
library('BSgenome')
```


## old vs new

In terms of installing BioConductor genomes and libraries they have recently deprecated some functions

deprecated
```{r}
source("https://bioconductor.org/biocLite.R") # deprecated
biocLite('BSgenome') # deprecated
```

New!

```{r}
BiocManager::install("BSgenome") 
```

You only need one, and depending on what version you have, it should be easy to figure out, however, you will want to migrate to the newer version soon.

### Downloading the Drosophila genome

First let us take a look at available genomes

```{r}
available.genomes()
```

This is of course only a handful of genomes to play with, and getting your own genome of interest installed is something for you to do. We with work with a recent version of the Drosophila melanogaster genome assembly

```{r}
#biocLite("BSgenome.Dmelanogaster.UCSC.dm6")
BiocManager::install("BSgenome.Dmelanogaster.UCSC.dm6")

library(BSgenome.Dmelanogaster.UCSC.dm6)
dm <- BSgenome.Dmelanogaster.UCSC.dm6
```


## Some basic statistics about the genome
```{r}
length(dm)
names(dm)
show(dm)
organism(dm)
seqinfo(dm)
isS4(dm)
seqlengths(BSgenome.Dmelanogaster.UCSC.dm6)
alphabetFrequency(dm$chr2R, 
                  baseOnly = T, as.prob = T)
```


```{r}
seqinfo(dm)
sum(seqlengths(dm))/1e6


dm_chrX <- getSeq(dm, "chrX")
nchar(dm_chrX)/1e6
```

## Frequencies of nucleotides on the X chromosome
```{r}
alphabetFrequency(dm_chrX, baseOnly = T, as.prob = T)

```


## Find some simple patterns
```{r}
seqs <- c("CCCGGG", "CCCCTTTT", "AAATTT", "TAGCT")
n = sapply(seqs, function(x) countPattern(x, dm_chrX ))
n
which.max(n)
```

## patterns

In BioStrings, there are a few functions

  - `matchPattern(pattern, subject, max.mismatch = 1)` one string to one string
  - `vmatchPattern(pattern, subject)`
    - one set of strings to one string
    - one string to one set of strings.
  - `findPalindromes` - finds palindromic regions in a single string.
 
 Other tools for PWMs, dictionaries etc..(?`vcountPDict,BSgenome-method`)
 - see `?PWM` and also `matchLRPatterns`


## a position weight matrix for a Drosophila transcription factor HNF4alpha

THis is a TF that is already in bioconductor (although you may have to install)
```{r}
data(HNF4alpha)
HNF4alpha
class(HNF4alpha)
```


We can generate a position frequency matrix and position weight matrix of these binding sites.
```{r}
pfm <- consensusMatrix(HNF4alpha)
round(pfm[1:4,]/colSums(pfm[1:4,]), 2)
class(pfm)

pwm <- PWM(pfm)

# or directly
pwm <- PWM(HNF4alpha, type = "prob")
round(pwm[1:4,]/colSums(pwm[1:4,]), 2)
maxWeights(pwm)

conMat <- consensusMatrix(HNF4alpha, as.prob = T)[1:4,]
seqLogo(conMat)
seqLogo(pwm[1:4,]/colSums(pwm[1:4,]))
```

## matching these to the X chromosome

```{r}
TF_hits <- matchPWM(pwm, dm_chrX, with.score = TRUE)
length(TF_hits)
head(mcols(TF_hits)$score)

hist(mcols(TF_hits)$score)
```

## How about the reverse complement of the X?

```{r}
TF_hits_rc <- matchPWM(reverseComplement(pwm), dm_chrX, with.score = TRUE)
length(TF_hits_rc)
hist(mcols(TF_hits_rc)$score)
```

## Extracting upstream sequences from Drosphila melanogaster

Makes a TxDb object (transcript annotations. Again, lots of options depending on how you want to refer to genes. Then extract the upstream sequences from the genes.
```{r}
dm_txdb <- makeTxDbFromUCSC("dm6", tablename="refGene") 

up10000seqs <- extractUpstreamSeqs(dm, dm_txdb, width=10000)
```

How about this pattern?

```{r}
upstream_hits <- vcountPattern("NGNNCAAAGNNCA", up10000seqs)
length(upstream_hits)
```


