---
title: "Control flow in R, the fundamentals"
author: "Ian Dworkin"
date: "`r format(Sys.time(),'%d %b %Y')`"
output: 
  html_document: 
    toc: yes
    number_sections: yes
    keep_md: yes
editor_options: 
  chunk_output_type: console
---


# Worked examples using control flow

```{r}
knitr::opts_chunk$set(eval = FALSE)
```

**NOTE:** I have set the defaults (line above) for evaluation of blocks of code to be off (`eval = FALSE`). When you are running the code you might need to change these. I did this for rendering this activity.


## if else

Here is an example of using `if` and `else` to do different computation (or in this case, different output) depending on the input

```{r}
p_test <- function(p){
    if (p <= 0.05) print ("yeah!!!!") 
        else if (p >= 0.9) print ("high!!!!")
        else print ("somewhere in the middle")
}
```


Try some numbers
```{r}
p_test()
```


Try a random number from a uniform distribution on $[0,1]$ using `runif`

```{r}
(test_val_1 <- runif(1))
p_test(test_val_1)
```


But we often don't have just a single number, but a vector of numbers we need to evaluate and make decisions with. 

```{r}

```

