---
title: "In class activity. A worked example using control flow (for loops, if statements, etc)"
author: "Ian Dworkin"
date: "`r format(Sys.time(),'%d %b %Y')`"
output: 
  html_document: 
    toc: yes
    number_sections: yes
    keep_md: yes
editor_options: 
  chunk_output_type: console
---


# Introduction

Let's do a little exercise integrating some of the things we have learned. Here are some Illumina HiSeq reads for one of our recent projects:

```{r}
read_1 <- "CGCGCAGTAGGGCACATGCCAGGTGTCCGCCACTTGGTGGGCACACAGCCGATGACGAACGGGCTCCTTGACTATAATCTGACCCGTTTGCGTTTGGGTGACCAGGGAGAACTGGTGCTCCTGC"

read_2 <- "AAAAAGCCAACCGAGAAATCCGCCAAGCCTGGCGACAAGAAGCCAGAGCAGAAGAAGACTGCTGCGGCTCCCGCTGCCGGCAAGAAGGAGGCTGCTCCCTCGGCTGCCAAGCCAGCTGCCGCTG"

read_3  <- "CAGCACGGACTGGGGCTTCTTGCCGGCGAGGACCTTCTTCTTGGCATCCTTGCTCTTGGCCTTGGCGGCCGCGGTCGTCTTTACGGCCGCGGGCTTCTTGGCAGCAGCACCGGCGGTCGCTGGC"
```

Question 1. what species are these sequences from?

Question 2. Put all three of these reads into a single object (a vector) called "reads".  What class will the vector `reads` be? Check to make sure! How many characters are in each read (and why does `length()` not give you what you want.. try...)

Question 3. Say we wanted to print each character (not the full string) from read_1, how do we do this using a for loop? You may wish to look at a function like `strsplit()` to accomplish this (there are other ways.)

Replace the blanks.
```{r}
read_1_split <- strsplit(___, split = ___, fixed = T) 
```

Question 4. What kind of object does this return? How might we make it a character vector again?

Question 5. How about if we wanted the number of occurrences of each base? Or better yet, their frequencies? You could write a loop, but I suggest looking at the help for the `table()` function... Also keep in mind that for for most objects `length()` tells you how many elements there are in a vector. For lists use `lengths()` (so you can either do this on a character vector or a list, your choice)

Question 6. How would you make this into a nice looking function that can work on either  a list or vectors of characters? (Still just for a single read)

Question 7. Now how can you modify your approach to do it for an arbitrary numbers of reads? You could use a loop or use one of the apply like functions (which one)?

Question 8. Can you revise your function so that it can handle the input of either a string as a single multicharacter vector, **or** a vector of individual characters **or** a list? Try it out with the vector of three sequence reads (`reads`).  
