---
title: "A worked example using control flow (for loops, if statements, etc)"
author: "Ian Dworkin"
output: 
  html_document: 
    keep_md: yes
    number_sections: yes
---

Let's do a little exercise integrating some of the things we have learned. Here are some Illumina HiSeq reads for one of our recent projects:

```{r}
read_1 <- "CGCGCAGTAGGGCACATGCCAGGTGTCCGCCACTTGGTGGGCACACAGCCGATGACGAACGGGCTCCTTGACTATAATCTGACCCGTTTGCGTTTGGGTGACCAGGGAGAACTGGTGCTCCTGC"

read_2 <- "AAAAAGCCAACCGAGAAATCCGCCAAGCCTGGCGACAAGAAGCCAGAGCAGAAGAAGACTGCTGCGGCTCCCGCTGCCGGCAAGAAGGAGGCTGCTCCCTCGGCTGCCAAGCCAGCTGCCGCTG"

read_3  <- "CAGCACGGACTGGGGCTTCTTGCCGGCGAGGACCTTCTTCTTGGCATCCTTGCTCTTGGCCTTGGCGGCCGCGGTCGTCTTTACGGCCGCGGGCTTCTTGGCAGCAGCACCGGCGGTCGCTGGC"
```

Question1: what species are these sequences from?

Let's put these reads into a single object ( a vector)
```{r}
reads <- c(read_1, read_2, read_3)
```
Question 2. What class will the vector `reads` be? Check to make sure!

Say we wanted to print each character from read_1, how do we do this using a for loop? You might try something like""
```{r}
for (i in read_1) {
	print(i)
}
```

While this prints out the full string, it does not do so character by character. THis does work in some other languages though.


We can make this one string into a vector of strings with each element being a single letter

```{r}
read_1_split <- strsplit(read_1, split = "", fixed = T) # a list
```

Since it is a list, which is annoying we want to coerce this into a character vector so we can work with it more easily.

```{r}
read_1_char <- as.character(unlist(read_1_split))
mode(read_1_char)
```

Printing the list once:

```{r}
for (i in read_1_split){
	print(i)
}
```

**VS** looping through each character

```{r}
for (i in read_1_char){
	print(i)
}
```

How about if we wanted the number of occurrences of each base? Or their frequencies? Well, to start with the easiest thing to do would be to use the `table()` function!

For a list you could do this:

```{r}
site_freq <- table(read_1_split)/lengths(read_1_split)
print(site_freq, digits = 2)
```
Note the use of `lengths()` not `length()` for the list above!!!!

For a vector of characters you can use this:

```{r}
site_freq_alt <- table(read_1_char)/length(read_1_char)
print(site_freq_alt, digits = 2)
```

Since this is a vector of characters we can use `length()`

## How would you make this into a nice looking function that can work on either lists or vectors of characters?

You could write a function like:

```{r}
BaseFrequencies <- function(x) {
    if (mode(x) == "list") {
    	tab <- table(x)/lengths(x)}
    
    else {
    	tab <- table(x)/length(x)
    }	
   return(tab)
}
```
Question 4. Add into the function what to do if it could also come in as a single multi-character string.


## Now try doing it with sapply across the three reads

It may be worth re-writing this function to be a bit more general.

```{r}
BaseFrequencies <- function(x) {
    
    # if it is a single string still
    if (length(x) == 1 & mode(x) == "character") {
    	x <- strsplit(x, split = "", fixed = T) 
        x <- as.character(unlist(x))
    }     
    
    if (mode(x) == "list") {
    	tab <- table(x)/lengths(x)}
    
    else {
    	tab <- table(x)/length(x)
    }	
   return(tab)
}
```
Let's try it out:

```{r}
basefreq <- sapply(reads, BaseFrequencies, USE.NAMES = F)
print(basefreq, digits = 2)
```