---
title: "MakingAndBreakingFunctions"
author: "Ian Dworkin"
date: "15/11/2021"
output: 
  html_document: 
    toc: yes
    number_sections: yes
    keep_md: yes
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(digits = 2)
```


## Why write a function in R?

- We have now used a few built in functions in R (there are many). 
- anything where you use "()" is a function.

**But why might we want to write our own functions?**


## Why write your own functions

However, if you write some good functions and scripts people will want to use them and edit them for their own purposes. Of course, they (and even you 6 months later) don't always know what the functions are always supposed to do or not do. So it is useful to always build robust functions that spit out warnings and errors, and also to consider unit tests.

As a review [here](https://r4ds.had.co.nz/functions.html#when-should-you-write-a-function) and [here](http://adv-r.had.co.nz/Functions.html)

## When should you write a function
Have you ever copy and pasted a block of code more than a couple of times? If so, write a function!


 Thankfully it is very easy to write our own functions in R. You should definitely get in the habit of doing so whenever you are going to need to do something more than once. You can collect all of these functions in a seperate script and can use source() to read them in.
 
 
## functions have the following format
```{r, eval = FALSE}
aFunction <- function(input variable 1, input variable 2, argument, etc...) {
expressions to calculate}
```

- This is abstract so let me give you a real example

We want to compute the standard error of the mean (SEM) which is approximately equal to the standard deviation divided by the square root of the sample size. How might we do it?

## standard error of the mean 

```{r}
a <- c(1,2,3,5,7,3,2,5,2)
sd_a <- sd(a)
sample_a <- length(a)

sd_a/sqrt(sample_a)
```


Or we could do it in one line. ** Notice the function within a function**

```{r}
sd(a)/sqrt(length(a)) 
```


But what happens if we have another function `b`

```{r}
b <- rnorm(100, mean = 10, sd = 1)

sd(b)/sqrt(length(b)) 

```

or c? d?

 This gets not only very monotonous, but very error prone!
## SEM as a function

 Try to write a function for the SEM 
 
 
## SEM function

```{r}
StdErr <- function(vector_vals) {
       sd(vector_vals)/sqrt(length(vector_vals)) } 
```


Now we can just use the variable `a` we created above that stores the vector of numbers and plug it in to our function. 

```{r}
StdErr(a)
```


We can also easily use it for other data like in the object `b`

```{r}
StdErr(b)
```


## taking a look at our function


Just type the name of the function to see what the function is.

```{r}
StdErr
```


And if you want to edit it

```{r}
edit(StdErr)
```


## looking at base R functions does not always work so easily

```{r}
mean
```


This is because many of the functions are either hidden or in this case written in another programming language that R calls. You can [read here](https://stackoverflow.com/questions/19226816/how-can-i-view-the-source-code-for-a-function)



## What if you want to return multiple things from your function


Practically speaking the last thing you write will be returned in a function, but it is better to be a bit formal about it with the `return` function.

So this will not behave as you might expect. Why?


```{r}
StdErr_V2 <- function(vector_vals) {
       se <- sd(vector_vals)/sqrt(length(vector_vals))} 
```


```{r}
StdErr_V2(a)
```


## using return

```{r}
StdErr_V3 <- function(vector_vals) {
       se <- sd(vector_vals)/sqrt(length(vector_vals))
       return(se)} 
```


```{r}
StdErr_V3(a)
```

## you can give the output names too

```{r}
StdErr_V4 <- function(vector_vals) {
       se <- sd(vector_vals)/sqrt(length(vector_vals))
       return(c(StandardError = se))} 
```

```{r}
StdErr_V4(a)
```


You may have noticed that despite only having one output, I have concatenated this. This is a bit of an R thing, but it comes in handy when we want multiple outputs


```{r}
StdErr_V5 <- function(vector_vals) {
       se <- sd(vector_vals)/sqrt(length(vector_vals))
       return(c(StandardError = se, 
                StandardDeviation = sd(vector_vals)))} 
```

```{r}
StdErr_V5(a)
```

## See if you can modify this to output a list instead of a vector and also output the mean



```{r}
StdErr_V6 <- function(vector_vals) {
       se <- sd(vector_vals)/sqrt(length(vector_vals))
       return(list(StandardError = se, 
                   StandardDeviation = sd(vector_vals), 
                   Mean_val = mean(vector_vals)))} 
```

```{r}
StdErr_V6(a)
```


## BREAK

## The function you can work with in groups

This function computers the [coefficient of variation.](https://en.wikipedia.org/wiki/Coefficient_of_variation), a positive, unitless measure of relative variation. It is nothing more than the ratio of the standard deviation to the mean. 


## Coefficient of variation

```{r}
CoefVar_v1 <- function(x) {
    cv <- sd(x)/mean(x)
}
```


## An example with a ratio

let's use this example with computing the ratio of two input variables


```{r}
ratio_function <- function(x, y) {
    x/y
}
```



```{r}
x <- CoefVar_v1(rnorm(100, 10, 3))
x
```

```{r}
ratio1 <- ratio_function(rnorm(100, 10, 3), rnorm(100, 10, 3))
ratio1
```

## Time to break it. 
In groups of 4, each of you will (one at a time) give your version of the function to one other person in the group, who will write some code to "break" the function in some way. Don't tell your neighbours what the variables are, just let them try and use it and break it...


## Things to throw into it

- negative values that give a -ve CV
- NA
- non integer or numeric
- vector of length zero

## Fixes
- `stop` with an if
- `break`
-`stopifnot`
- checking object class
- `stop` vs. `warning` vs. message
-`exist` or `file.exists()`


## 
```{r}
ratio_function_v2 <- function(x, y) {
    if (length(x) != length(y)) stop("x and y are not the same length")
    if (mode(x) != "numeric") stop("x is not numeric")
    if (mode(y) != "numeric") stop("y is not numeric")
    if (anyNA(x) || anyNA(y)) warning("missing data present")
    x/y
}

```

```{r}
x <- rnorm(100, 10, 2)
y <- rnorm(100, 50, 3)
#x <- sample(c("a", "b", "c"), size = 100, replace = T)
#y[45] <- NA
#any(is.na(y))

y <- 1:100
ratio_function_v2(x, y)

rm(x, y)
```

