---
title: "Bio720 Introduction to `R`, in class exercise"
author: "Ian Dworkin"
date: "October 3, 2016"
output: html_document
---
## Overview
In class tonight we are going to both practice some of the `R` skills you were introduced to in the video tutorials. [Click here](https://github.com/DworkinLab/Bio720/blob/master/Introduction_to_R.md) for that list.

The learning objectives for today are as follows:
1. Learn some best practices for organizing computational projects (i.e. any projects with some scripts, data and outputs).
2. Learn some intuitive (but not necessarily technical) ideas about *data structures* in general, and review some of the data structures in `R`.
3. Practice some of the skills that were introduced in the video tutorials.

## How to organize computational projects.

Please [click here](https://github.com/DworkinLab/Bio720/blob/master/IntroductionMarkdownAndVersionControl/Bio720_IntroductionMarkdown.md#a-few-words-on-project-organization) to link to a brief discussion on these points.

## Some very basic thoughts on *data structures* in `R`
We are not going to have a computer science-esque discussion of data structures (there are whole courses on this), but instead try to introduce a few basic concepts to understand why computers need to store different types of data in different ways, and why we need to be aware of that.

### What is the point of data structures? (class discussion)
- What kind of data do we want the computer to store for us?
- Why does it matter what kind of data (integers, floating point numbers, strings, Boolean,...)?

### Data structures in R

As was discussed in the video screencasts R has a number of different basic data structures (and more can be made or extended by programmers like you!). We need to start with the so-called *atomic* types that can be stored as vectors (remember R does not have an object to store scalars!). You can google them, but they are logical (Boolean), integer, real (double or floating point) , complex, string (or character) and raw. Let's think about a few of them:



```{r}
x <- 1
```

and you can find out information about this with a variety of functions:

```{r}
str(x)
mode(x)
typeof(x)
```
Why does `mode(x)` and `typeof(x)` give different results?


Let's create a few more objects and take a look at them

```{r}
y = c(3, 4, 5)
```

Will `x` and `y` differ?  Check and explain?

Now let's create a new object z:

```{r}
z = 3:5
```
How should `y` and `z` compare? how would you check?

```{r}
mode(y)
mode(z)
typeof(y)
typeof(z)
```
 Which might suggest they are different. In one case `R` is treating the vector as integers, the other case as floating point (double). So this suggests they might be different. However:
 
```{r}
y == z
all.equal(y, z)
# BUT
identical(y, z)
```

Suggests R is treating them the same. This is definitely one of the odd R behaviours. While in many languages (like `C++`) you need to define the type of variable you are creating, R tries to make "guesses" about what you are doing. Sometimes this can result in odd behaviour.


Ok, let's think about some of the other basic data types we learned about (strings or "character" in R, boolean/logical)

```{r}
rm( x, y, z) # clean up
x = 1
y = "1"
z = "one"
a = TRUE
b = "TRUE"
```

Before checking, think about what types each of these objects should be. Then check.
```{r}
x == y
all.equal(x, y)
identical(x, y)
```

How about `y` and `z`?

```{r}
y == z
all.equal(y, z)
identical(y, z)

# Which might be obvious by
mode(y)
mode(z)
```


```{r}
y == z
all.equal(y, z)
identical(y, z)

# Which might be obvious by
mode(y)
mode(z)
```

The same "logic" applies for comparing `a` and `b`

```{r}
a == b
all.equal(a, b)
identical(a, b)

# Which might be obvious
mode(a); typeof(a)
mode(b); typeof(b)
```

So, what are the take home points?
- Computers can be very exact.
- R has some confusing ideas about storing variables, assigning types to basic variables and you need to think carefully about what "equals" means.

Group exercises: How would you get `R` to coerce `x` and `y` to be exactly the same? How about `a` and `b`? 

Let's coerce b into a logical/Boolean
```{r}
b1 <- as.logical(b)
b == b1
a == b
typeof(b1); mode(b1)
```

Let's coerce a into a character
```{r}
a1 <- as.character(a)
a == a1
a == b
typeof(a1); mode(a1)
```

And, y into a number
```{r}
y1 <- as.numeric(y)
y == y1
identical(y, y1)

x == y1
identical(x, y1)
typeof(y1); mode(y1)
```


You can also embed plots, for example:

# # ```{r, echo=FALSE}
# plot(cars)
# ```

