---
title: "Bio720 Introduction to `R`, in class exercise"
author: "Ian Dworkin"
date: "October 3, 2016"
output: html_document
---
## Overview
In class tonight we are going to both practice some of the `R` skills you were introduced to in the video tutorials. [Click here](https://github.com/DworkinLab/Bio720/blob/master/Introduction_to_R.md) for that list.

The learning objectives for today are as follows:
1. Learn some best practices for organizing computational projects (i.e. any projects with some scripts, data and outputs).
2. Learn some intuitive (but not necessarily technical) ideas about *data structures* in general, and review some of the data structures in `R`.
3. Practice some of the skills that were introduced in the video tutorials.

## How to organize computational projects.

Please [click here](https://github.com/DworkinLab/Bio720/blob/master/IntroductionMarkdownAndVersionControl/Bio720_IntroductionMarkdown.md#a-few-words-on-project-organization) to link to a brief discussion on these points.

## Some very basic thoughts on *data structures* in `R`
We are not going to have a computer science-esque discussion of data structures (there are whole courses on this), but instead try to introduce a few basic concepts to understand why computers need to store different types of data in different ways, and why we need to be aware of that.

### What is the point of data structures? (class discussion)
- What kind of data do we want the computer to store for us?
- Why does it matter what kind of data (integers, floating point numbers, strings, Boolean,...)?

### Data structures in R

As was discussed in the video screencasts R has a number of different basic data structures (and more can be made or extended by programmers like you!). We need to start with the so-called *atomic* types that can be stored as vectors (remember R does not have an object to store scalars!). You can google them, but they are logical (Boolean), integer, real (double or floating point) , complex, string (or character) and raw. Let's think about a few of them:

```{r}
x <- 1
```

and you can find out information about this with a variety of functions:

```{r}
str(x)
mode(x)
typeof(x)
```
Why does `mode(x)` and `typeof(x)` give different results?


Let's create a few more objects and take a look at them

```{r}
y = c(3, 4, 5)
```

Will `x` and `y` differ?  Check and explain?

Now let's create a new object z:

```{r}
z = 3:5
```
How should `y` and `z` compare? how would you check?

Ok, let's think about some of the other basic data types we learned about (strings or "character" in R, boolean/logical)

```{r}
rm( x, y, z) # clean up
x = 1
y = "1"
z = "one"
a = TRUE
b = "TRUE"
```

Before checking, think about what types each of these objects should be. Then check.

How about `y` and `z`? `a` and `b`?


```{r}

```

So what are the take home points of these demonstrations?

Group exercises: How would you get `R` to coerce `x` and `y` to be exactly the same? How about `a` and `b`? 


## Building up our data structures. 

Now that we have some better idea (hopefully) of some of the atomic data types, we want to use these to build more complex data structures that may (eventually - like next week) be useful for data analysis, simulations and the like. There are a few important ones that we will use a lot: matrix, list, data.frame, factors, and formula (which we will not cover in Bio720 but is essential for statistical analyses). There are other important ones (like array) but we will cover these other ones first.

Before we get any further and create some new objects, how do we see all of the objects we currently have in our global environment?

```{r}
ls()
```

Let's work with a clean slate. How might we remove all of the objects and start fresh? Obviously you could just do a `rm()` command with each object name, but you can also remove all at once.

```{r}
rm(list=ls())
ls()
```

Let's think about what this command has done.



You can also embed plots, for example:

# # ```{r, echo=FALSE}
# plot(cars)
# ```

