---
title: "Bio720_R_Assignment_3_Answers"
author: "Ian Dworkin"
date: "11/7/2018"
output: 
  html_document: 
    keep_md: yes
    number_sections: yes
    toc: yes
---
# Answers to *R* asssignment 3
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)



```

## Question 2
Or wherever you have downloaded the data
```{r}
rna_counts <- read.csv("eXpress_dm_counts.csv", h = T)
```

Write the function

```{r}
mean_function <- function(x , log2Tr = FALSE) {
	if (log2Tr == TRUE) {
	  x <- log2(x + 0.0001)}
	mean(x)
}
```

Check that it works
```{r}
mean_function(rna_counts[,2])
mean(rna_counts[,2])
mean_function(rna_counts[,2], TRUE)
mean(log2(rna_counts[,2]+ 0.0001)) # same
```

## Question 3
The -1 in `rna_counts[,-1]` is to just exclude the first column filled with identifiers.

```{r}
stored_values <- rep(NA, ncol(rna_counts[,-1]))
length(stored_values)

for (i in 1:ncol(rna_counts[,-1])) {
	stored_values[i] <- mean_function(rna_counts[,i+1], TRUE)
}

names(stored_values) <- colnames(rna_counts[,-1])
stored_values
length(stored_values)
```

Always double check your results

```{r}
stored_values[55]
mean(log2(rna_counts[,56]+ 0.0001))
```

## Question 4

```{r}
apply(rna_counts[,-1], MARGIN = 2, mean_function, TRUE)
```
You could have used sapply or tapply as well.

## Question 5

```{r}
colMeans(log2(rna_counts[, -1] + 0.0001))
colMeans(rna_counts[, -1])
```

## Question 6
It is common (say for a MAplot) to want the mean expression value of each given gene across all samples. Write a function to do this, and using one of the approaches from Q 3-5 generate and store these values in a variable.

I used the same function I wrote above, and just switched the apply function from using the function per column to per row.
```{r}
gene_means <- apply(rna_counts[,-1], MARGIN = 1, mean_function, TRUE)
nrow(rna_counts)
length(gene_means)
```

Even easier to use `rowMeans()`. Below that is just demonstrating that you get the same answer.

```{r}
gene_means_alt <- rowMeans(log2(rna_counts[, -1] + 0.0001))

length(gene_means_alt)
head(gene_means)
head(gene_means_alt)
```

## Question 7
We are very interested in what is going on in the head horns between small males and large males. Using the type of tools you have written (feel free to modify as you need, but show the new functions) calculate the mean expression for the subset of columns for large and small male head horns. Note you are calculating means on a gene by gene basis, NOT sample by sample. Now calculate the mean difference (again gene by gene) between large male and small males (for head horns). i.e. first calculate the mean expression among individuals who are large males (head horns), ditto for the small males, and calculate their difference.

### Answer

First let's look at the names of the columns
```{r}
names(rna_counts)
#colnames(rna_counts) also works
```

We could do this manually and just extract the columns, but it is better to do this in an automated fashion. We can use a simple regular expression to help here using the `grep` family of functions, in particular the logical form of grep, `grepl()`

```{r}
grepl("male_hdhorn", names(rna_counts))
```

So we set this up for both small male and large male.

```{r}
sm_m_hd <- grepl("sm_male_hdhorn", names(rna_counts))

lg_m_hd <- grepl("lg_male_hdhorn", names(rna_counts))
sum(sm_m_hd)
sum(lg_m_hd)
```

Confirming 4 samples of each. Now we can extract the columns we want.

```{r}
sm_m_hd_dat <- rna_counts[, sm_m_hd]
lg_m_hd_dat <- rna_counts[, lg_m_hd]

dim(sm_m_hd_dat)
dim(lg_m_hd_dat)

head(sm_m_hd_dat)
```

Now we can go ahead and answer the question

```{r}
mean_hd_horn_expression <- rowMeans(log2(cbind(sm_m_hd_dat, lg_m_hd_dat) + 0.1))

length(mean_hd_horn_expression)

mean_lg_hd_horn_expression <- rowMeans(log2(lg_m_hd_dat + 0.1))

mean_sm_hd_horn_expression <- rowMeans(log2(sm_m_hd_dat + 0.1))

hd_horn_diff <- mean_lg_hd_horn_expression - mean_sm_hd_horn_expression
```


## Question 8

For the log transformed data

```{r}
plot(y = hd_horn_diff, x = mean_hd_horn_expression,
     pch = 20, col = "#fe000032",
     ylab = "log2 Difference of Large VS. Small head horns",
     xlab = "log2 Mean expression by gene")
```

Note what happens if you add a smaller amount to offset zero counts... THis is something to be wary of, and is handled very differently in real analyses!
